# Analyse gender in music psychology

T. Eerola, 19/3/2025

Gender analysis of the WEIRD data (Jakubowski et al., 2025).

## Required libraries

```{r}
#| id: libraries
#| warning: false
#| eval: true
#| echo: false
#| message: false
library(tidyverse)
library(stringr)
library(ggplot2)
library(ggwordcloud)
library(modelsummary)
```


## Load and preprocess data

```{r}
#| id: preprocess
#| eval: true

full_process <- FALSE
if(full_process==TRUE){
    d <- read.csv('../data/WEIRD part cleaned Feb24.csv',header = TRUE)
    source('scripts/preprocess.R')      # Routine to process the WEIRD dataset
    source('scripts/gender_analysis.R') # creates enriched dataset
}
#### Get processed data --------------------
if(full_process==FALSE){
    d <- readRDS('../data/enriched_dataset.rds')
}

D <- dplyr::filter(d,study_id=='study1') # 1360 articles
print(paste("number of unique studies:",nrow(D)))
```

## Deal with other authors (non-first authors)

This needs to be done. 

1. The author first names for all authors (except for the 1st author) are not in the WEIRD data frame, so these needs to be retrieved.   

2. Then the gender attribution process needs to be done with some manual checking.

3. Aggregate with the present (first author) data 

## Summarise

This is now only first author data.

```{r}
#| id: summarise
#| eval: true
#| output: asis
source('scripts/summarise_gender.R') #
```

## Visualise

This is now only first author data.

```{r}
#| id: visualise
#| eval: true
source('scripts/visualise_gender.R') # creates a figure
source('scripts/visualise_gender_trends.R') # creates a figure
source('scripts/visualise_names.R')  # creates a figure
```
